---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
library(inferSEM)
library(mxsem)
```

# inferSEM

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![R-CMD-check](https://github.com/jhorzek/inferSEM/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/jhorzek/inferSEM/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

inferSEM is an experimental R package for inference on structural equation models
fitted with OpenMx. The objective is to apply the same ideas used in Bayesian networks
for inference in SEM.

inferSEM currently only has one user facing function: `infer`. This function allows
computing the interventional and conditional distributions of SEMs. However,
it is not really tested yet and the implementation could be wrong. This is just a 
small experiment package. For a thorough introduction to interventional and 
conditional distributions in SEM, please see Gische et al. (2021).

The interventional distribution is computed as follows: Say we intervene by setting
the value of the variable x2 to 1. In that case:

1. The RAM matrices are extracted from the SEM
2. The SEM is "mutilated" - a term used in Bayesian networks (see Scutari, 2010) to describe the removal
of any effects on the intervention variable (x2 in our case). In SEM this also includes 
the removal of covariances between x2 and any other variable in the model
3. The computation of the model implied means and covariances based on the mutilated SEM

The conditional distribution is computed using the R package condMVNorm. If interventional
and conditional distribution are combined, inferSEM first computes the interventional
distribution and then conditions on the additional evidence.

## Installation

You can install the development version of inferSEM as follows:

``` r
if(!require("remotes"))
install.packages("remotes")
remotes::install_github("jhorzek/inferSEM")
```

## Example

In the following, we will use the R package mxsem to create the OpenMx model.

```{r example}
library(inferSEM)
library(mxsem)
# The following model is based on the documentation of sem in lavaan:
model <- ' 
  # latent variable definitions
     ind60 =~ x1 + x2 + x3
     dem60 =~ y1 + a*y2 + b*y3 + c*y4
     dem65 =~ y5 + a*y6 + b*y7 + c*y8

  # regressions
    dem60 ~ ind60
    dem65 ~ ind60 + dem60

  # residual correlations
    y1 ~~ y5
    y2 ~~ y4 + y6
    y3 ~~ y7
    y4 ~~ y8
    y6 ~~ y8
'

fit <- mxsem(model = model,
             data  = OpenMx::Bollen) |>
  mxTryHard()
```

First, let's have a look at the unconditional means and covariances of the latent
variables:
```{r}
unconditional <- inferSEM::infer(model = fit)
# For simplicity, let's just look at the latent variables:
round(unconditional$means[,c("ind60", "dem60", "dem65")], 2)
round(unconditional$covariances[c("ind60", "dem60", "dem65"),
                                c("ind60", "dem60", "dem65")], 2)
```

Now, let's assume we are interested in the distribution of our variables given
that we intervene on dem60 with do(dem60 = 3). We can get the interventional distribution
as follows:

```{r}
interventional <- inferSEM::infer(model = fit,
                                  intervene = c("dem60" = 3))
# For simplicity, let's just look at the latent variables:
round(interventional$means[,c("ind60", "dem60", "dem65")], 2)
round(interventional$covariances[c("ind60", "dem60", "dem65"),
                                 c("ind60", "dem60", "dem65")], 2)
```
In contrast, if we just observed dem60 with observe(dem60 = 3), we get:

```{r}
conditional <- inferSEM::infer(model = fit,
                               observe = c("dem60" = 3))
# For simplicity, let's just look at the latent variables:
round(conditional$means[,c("ind60", "dem65")], 2)
round(conditional$covariances[c("ind60", "dem65"),
                              c("ind60", "dem65")], 2)
```
Note that dem60 is no longer included in the output because we are conditioning on
it.

We can also plot the distributions:

```{r}
x <- seq(0, 5, by = .01)
interventional_dem65 <- dnorm(x = x,
                              mean = interventional$means[,"dem65"],
                              sd = sqrt(interventional$covariances["dem65", "dem65"]))
conditional_dem65 <- dnorm(x = x,
                           mean = conditional$means[,"dem65"],
                           sd = sqrt(conditional$covariances["dem65", "dem65"]))
plot(x = x, 
     y = interventional_dem65, 
     type = "l",
     col = "red",
     xlab = "dem65",
     ylab = "Density")
lines(x = x, 
      y = conditional_dem65,
      col = "blue")
```


## References

- Gische, C., West, S. G., & Voelkle, M. C. (2021). Forecasting causal effects
of interventions versus predicting future outcomes. Structural Equation Modeling:
A Multidisciplinary Journal, 28(3), 475-492.
- Scutari, M. (2010). Learning Bayesian Networks with the bnlearn R Package. Journal of Statistical Software,
35(3), 1-22. URL http://www.jstatsoft.org/v35/i03/.
